---
title: "Swiss Cantons"
author: "Marco Giordano"
date: "12/24/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

What if you don't find the plots you wish for online? Well, you actually have to make your own! I had this problem for a project where I had to gather some data by myself and show my findings. There were a lot of plots available but no one suited what I wanted to communicate. For this reason, I extracted data from the official Swiss Confederation website and cleaned them where necessary. After that, it was just a matter of transformations and plotting.

I included data in the "Data" folder so be sure to take a look at it if you want to replicate the result. You can find the link to the interactive tables below, wherein you can select what data you wish to include and then download them. 

Source: [Swiss Confederation website](https://www.pxweb.bfs.admin.ch/pxweb/en/px-x-0102020000_401/px-x-0102020000_401/px-x-0102020000_401.px)

```{r}
library(tidyverse)
library(ggthemes)
library(scales)
```

We load the raw data into R, considering that the encoding must be "latin1", otherwise you are likely to get problems while visualizing umlauts!

```{r, include=FALSE} 
swiss_data <- read.csv('Swiss_data_permanent_residents_def.csv', stringsAsFactors = F, encoding="latin1")
```

After we load the raw data, we have to process them since they are unusable in this format. e have to split a column into multiple, using the semicolon as a separator and then converting each new column to the right datatype.

```{r}
swiss_data <- separate(swiss_data, col = 'Permanent.resident.population.in.private.households.by.institutional.units.and.household.size', into = c('Canton', 'Household_Size', 'Resident_Population'), sep = ';')
swiss_data <- swiss_data[-c(1,7),]
swiss_prova <- swiss_data[str_detect(swiss_data$Canton, '^-'),]
swiss_prova$Canton <- str_replace_all(swiss_prova$Canton, '^-', ' ')
swiss_prova$Resident_Population <- as.integer(swiss_prova$Resident_Population)
swiss_prova$Household_Size <- as.factor(swiss_prova$Household_Size)
```

Now we are ready to apply some transformations to the data and then plot. Our goal is to find the top 5 Cantons by resident population, so we group by canton first and then calculate the total resident population for each of them and then sort. 

```{r}
swiss_data_top <- swiss_prova %>%
  group_by(Canton)%>%
  summarise(tot_pop = sum(Resident_Population)) %>%
  arrange(desc(tot_pop)) %>%
  ungroup() %>%
  top_n(5, tot_pop)
```

With our ordered data, we are only left with the plotting part. A barplot is one of the simplest but also most effective ways to show something. Originally, I had ordered data to find the top 10 Cantons but due to problems with the markdown, I have opted to pick 5, the concept is still the same!

```{r}
ggplot(swiss_data_top, aes(reorder(Canton, tot_pop), tot_pop, fill = 'red')) +
  geom_col() +
  theme_economist() +
  scale_color_economist() +
  theme(axis.title=element_text(size=14,face="bold")) +
  scale_y_continuous(labels = scales::comma) +
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle('Top 10 Swiss Cantons by Population') +
  xlab('Canton') +
  ylab('Resident Population (2019)') +
  guides(fill = FALSE)
```

As you can see above, 4 out of 5 are Swiss German Cantons, where Zürich naturally comes at the top as many could suspect. It is interesting to notice that Vaud (French Switzerland) is the only one in top 5 which is not German-speaking. We can clearly see the discrepancy between the first and the second place, with around half a million of difference in terms of resident population.

Now that we know which are the top Cantons, we are interested in knowing the composition according to the household size, we have to transform starting data differently.

```{r}
swiss_data_house <- swiss_prova %>%
  group_by(Canton) %>%
  mutate(tot_pop = sum(Resident_Population)) %>%
  ungroup() %>%
  top_n(30, tot_pop) %>%
  group_by(Canton, Household_Size) %>%
  arrange(desc(Resident_Population)) %>%
  ungroup()
```

Once we do that, we are ready to plot what we want by using a stacked barplot where one can visualize the distribution of the household size by Canton.

```{r}
ggplot(swiss_data_house, aes(reorder(Canton, Resident_Population), Resident_Population, fill = Household_Size)) +
  geom_bar(stat = 'identity', position = 'fill') +
  theme_economist() +
  scale_color_economist() +
  theme(axis.title=element_text(size=12,face="bold")) +
  scale_y_continuous(labels = scales::comma) +
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  ggtitle('Household Size for the top 5 Cantons') +
  xlab('Canton') +
  ylab('Resident Population (2019)') +
  guides(fill=guide_legend(title="Household Size"))
```

It is easy to see that there is a pretty similar distribution of household sizes across the top 5 Cantons, although one can look at which one is interested in. Consider also that around 50% of the population across the top Cantons is split into households of 3 or more persons.



